---
title: "Blast"
output: html_document
date: "2023-11-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Blast Tutorial Using Sample Geoduck File

Following the blast tutorial outlined in [TUSK](https://robertslab.github.io/tusk/modules/04-blast.html) to blast and annotate geoduck genes in the file provided in [this github issue](https://github.com/RobertsLab/resources/issues/1710). Important to note that the geoduck file is a *protein* file, so need to run *blastp* instead of *blastx* as listed in the tutorial.

Before starting, make a directory/repo to do all this in that contains the following folders:

-   Code (this is where this/the code file should be)

-   Data

-   Output

There are some packages that are used in this process including Dplyr, Stringr, ggplot2, and DT. Packages are installed/loaded here (disregard chunk if already loaded)

```{r}

# if you need to install, uncomment the following: 
# install.packages("BiocManager")
# BiocManager::install("Biostrings")
# install.packages("ggplot2")
# install.packages('DT')
# install.packages('dplyr')
# install.packages('stringr')

# load installed packages
library(BiocManager)
library(Biostrings)
library(ggplot2)
library(DT)
library(dplyr)
library(stringr)


```

## 1. Database Creation

[**This part is not unique to the file**]{.underline}, it is creating the database you will use when running blast on your file (ie. you don't need to change this section with each different file of interest).

### Obtain Fasta (UniProt/Swiss-Prot)

```{bash}
cd ../data
curl -O https://ftp.uniprot.org/pub/databases/uniprot/current_release/knowledgebase/complete/uniprot_sprot.fasta.gz
mv uniprot_sprot.fasta.gz uniprot_sprot_r2023_04.fasta.gz
gunzip -k uniprot_sprot_r2023_04.fasta.gz
```

### Making the Database

```{bash}
mkdir ../blastdb
/home/shared/ncbi-blast-2.11.0+/bin/makeblastdb \
-in ../data/uniprot_sprot_r2023_04.fasta \
-dbtype prot \
-out ../blastdb/uniprot_sprot_r2023_04
```

## 2. Getting the query fasta file

This is where you start changing things for your specific file of interest.

We already have our query rna fasta file.

Taking a peek at that file via `head()` and getting a count of sequences

```{bash}

head -3 ../data/ncbi_dataset/data/GCF_026571515.1/rna.fna

```

```{bash}

echo "How many sequences are there?"
grep -c ">" ../data/ncbi_dataset/data/GCF_026571515.1/rna.fna

```

## 3. Running Blastx

```{bash}
/home/shared/ncbi-blast-2.11.0+/bin/blastx \
-query ../data/ncbi_dataset/data/GCF_026571515.1/rna.fna \
-db ../blastdb/uniprot_sprot_r2023_04 \
-out ../output/GCF_026571515.1.tab \
-evalue 1E-20 \
-num_threads 20 \
-max_target_seqs 1 \
-outfmt 6
```

Peeking at the output file

```{bash}
head -2 ../output/GCF_026571515.1.tab
```

```{bash}
echo "Number of lines in output"
wc -l ../output/GCF_026571515.1.tab
```

## 
