---
title: "post filtering analysis"
author: "Megan Ewing"
date: "2024-07-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

|                                                                                                    |                                              |
|----------------------------------------------------------------------------------------------------|----------------------------------------------|
| **by group, files of interest**                                                                    |                                              |
| DEG stats                                                                                          | ../output/0725-DEGstats_filteredbygroup.tab  |
| DEG counts                                                                                         | ../output/0725-logcounts_allDEG_filtered.csv |
| Annot info                                                                                         | ../output/deg_blast_cds_full_0709.tab        |
|                                                                                                    |                                              |
| want to take all of those, and get the data for the list of genes that have 4 hits for pre filters |                                              |

### Reading in files of interest

*Prefilter paramaters: Average of 10+ reads across at least one treatment or control*

```{r}

DEGstats <- read.csv("../output/0725-DEGstats_filteredbygroup.tab", sep="")

DEGcounts <- read.csv("../output/0725-logcounts_allDEG_filtered.csv")

annot_info <- read.csv("../output/deg_blast_cds_full_0709.tab", sep="")

```
