---
title: "The Code That Never Was"
output: html_document
date: "2023-12-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The Code That Never Was

This is code that did not make the "final cut" and would've been nice, but ultimately not neccessary to the analyses.

## 06-deseq2

### Making an extra Heatmap for sig DEGs

I wanted to create a heatmap that showed sig. DEGs by sample, not neccessary, but was would've been cool. Scrapped it for the sake of time. The following code goes after the whole of the \`06-deseq2.Rmd\` script:

\~\~\~\~

commenting out this section so I can knit and folks can see the outputs

```{r}
# make a heatmap using only our significant DEGs

# #import data
# DEGCounts <- read.table("../output/1213-DEG.tab", header=TRUE, quote="\"")
# head(DEGCounts)
# 
# #select subset of just gene and sample columns
# DEGHeatmap <- DEGCounts[-c(2:7)]
# head(DEGHeatmap)
# 
# # establish that the first column is row names 
# # rownames(DEGHeatmap) <- DEGHeatmap$gene
# 
# #get rid of the redundant row names column that will be created.
# # DEGHeatmap <- DEGHeatmap[,-1]
# 
# 
# ### stuck here
# ### Error in seq.default(-m, m, length.out = n + 1) :'from' must be a finite number
# 
# #make heatmap
# pheatmap(DEGHeatmap, scale = "row",show_rownames = TRUE, treeheight_col = 1, fontsize_col = 10, fontsize_row = 3)



```

Chunk above is giving me trouble. Here's what I want:

1.  a data frame containing rows = genes, cols = samples, values = mean exp but ONLY for things where p \<= 0.05
2.  a heatmap that then shows the genes that are *significantly deferentially* expressed

```{r}

# read in sig.DEGs
DEGCounts <- read.table("../output/1213-DEG.tab", header=TRUE, quote="\"")
head(DEGCounts)

DEGCountNames <-rownames(DEGCounts)
DEGCounts <- add_column(DEGCounts, DEGCountNames, .before = DEGCounts$baseMean )

varGenes <- order( rowVars (x = assay(rld), useNames = FALSE ), decreasing=TRUE )
varGenes <-assay(rld)[topVarGenes,]

varNames <- rownames(varGenes)
varGenes <- as.data.frame(varGenes) 
varGenes <- add_column(varGenes, varNames, .before = varNames$`M-C-193`)

merge(varGenes, DEGCounts, by.y = "DEGCountNames", by.x = "top25names")

```
